---
title: "TheoEco HW 1"
author: "Dylan Sosa"
date: "1/17/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1

1. Discuss density-independent and density-dependent population growth. What is different about the growth rate in both cases? 
	1. Density-independent populations have exponential growth. They have a linear and fixed growth rate.
	2. Density-dependent populations have logistic growth and a characteristic sigmoid curve. They have a non-linear and non-fixed growth rate.
		
```{r results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(gridExtra)

exponential <-
  ggplot(data.frame(x = seq(0, 15)), aes(x)) +
  stat_function(fun = exp) +
  labs(title = 'Density Independent Pop Growth')

logistic <-
  ggplot(data.frame(x = seq(-15, 15)), aes(x)) +
  stat_function(fun = plogis) +
  labs(title = 'Density Dependent Pop  Growth')

grid.arrange(exponential, logistic, nrow = 1)
```

2. For population (or density, remember that for fixed area these are equivalent) dependent growth, explain positive and negative versions in terms of the dependence of the growth rate on population size. 
	1. The Allee effect is an example of positive density dependence of growth rate on population size. The rate $r$ is highest for small population sizes, $N$. These positive density dependent growth rate are unstable.
	2. As density increases over time resources become scarce, causing individuals to die off which gcauses negative density dependence. As the population $N$ increases, the growth rate $r$ decreases. 

3. Describe a brief biological or ecological scenario in which you might expect to find positive and negative density dependence. Keep it brief.
	1. Positive: Large animal aggregations like schools of fish have positive density-depenent effects. The population growth is regulated by the large number of prey which discourages predators. 
	2. Negative: Predation can have a negative density-dependent effect. When prey is at low density their population is left alone, but at high density the population can be brought to zero or near zero by predators.

# Problem 2
```{r results='hide', message=FALSE, warning=FALSE}
logGrowth <- function(x) {
  # r = 2 = u, c = 0.1, v = 5
  # r$N$(1-(N/K)) # logistic growth equation
  2 * x * (1 - (x / 15))
}

logGrowthPerCapita <- function(x) {
  (2 * x * (1 - (x / 15))) / x
}

alleeEffect <- function(x) {
  # r = 2 = u, c = 0.1, v = 5
  # ((u*N/v+N)-c*N)*N # Allee effect equation
  (((2 * x) / (5 + x)) - (0.1 * x)) * x
}

alleeEffectPerCapita <- function(x) {
  (((2 * x) / (5 + x)) - (0.1 * x))
}

plotPop <-
  ggplot(data.frame(x = seq(0, 15, 0.001)), aes(x)) +
  stat_function(fun = logGrowth, aes(color = 'log')) +
  stat_function(fun = alleeEffect, aes(color = 'Allee')) +
  labs(x = 'N', y = 'Population dN/dt', title = 'Logistic & Allee Effect Population Growth Rate') +
  scale_colour_manual("Legend", values = c("forestgreen", "brown")) +
  theme(
    plot.title = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )

plotPerCap <-
  ggplot(data.frame(x = seq(0, 15, 0.001)), aes(x)) +
  stat_function(fun = logGrowthPerCapita, aes(color = 'log')) +
  stat_function(fun = alleeEffectPerCapita, aes(color = 'Allee'), show.legend = TRUE) +
  labs(x = 'N', y = 'Per Capita dN/dt', title = 'Logistic & Allee Effect Per Capita Growth Rate') +
  scale_colour_manual("Legend", values = c("forestgreen", "brown")) +
  theme(
    plot.title = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )

grid.arrange(plotPop, plotPerCap, nrow = 1)
```

# Problem 3
```{r results='hide', message=FALSE, warning=FALSE}
## Bifurcation.R 
## Stefano Allesina sallesina@uchicago.edu allesinalab.uchicago.edu
## Code for "Theoretical Ecology"
## Modifications by Dylan

## This function returns the values of the min and max 
peaks <- function(x) {
  if (min(x)==max(x)) return(min(x)) ## Does not oscillate 
  l <- length(x) 
  xm1 <- c(x[-1], x[l])
  xp1 <- c(x[1], x[-l]) 
  z<-x[x > xm1 & x > xp1 | x < xm1 & x < xp1] 
  if (length(z)==0) return(min(x)) ## It has not converged yet 
  return (z)
} 

## This function creates a simulation of the logistic map 
LogisticMap<-function(N0,r,TimeSteps){
  Results<-rep(0,TimeSteps) 
  Results[1]<-N0 
  for (j in 2:TimeSteps){
    Results[j]<-r*Results[j-1]*(1-Results[j-1])
  } 
  return(Results)
} 

rickerMap<-function(N0,r,TimeSteps){
  Results<-rep(0,TimeSteps) 
  Results[1]<-N0 
  for (j in 2:TimeSteps){
    Results[j]<-exp(r*(1-(Results[j-1]/2)))*Results[j-1]
  } 
  return(Results)
}

## Plot the Diagram 
plot(0,0, xlim=c(0,4), ylim=c(-0.05,10),type="n", xlab="r", ylab="X") 
for (r in seq(0.001,100,0.005)) { # These are the initial and final values for r
  # out <- LogisticMap(0.5,r,2500) # Initial conditions 
  out <- rickerMap(0.5,r,2500) # Initial conditions 
  l <- length(out) %/% 10 # use only the last 250 steps 
  out <- out[(9*l):(10*l)] 
  p <- peaks(out) 
  l <- length(out) 
  points(rep(r, length(p)), p, pch=".")
}
```
Bifurcation diagram for the Ricker model map. Growth rate $r < 2$ is stable; $r == 2$ bifurcates into a 2-cycle; $r \approx 2.5$ bifurcates into a 4-cycle;  $r > 2.5$ seems to engender chaos while some regions (the white spots) return to lower cycles.

# Problem 4
```{r results='hide', message=FALSE, warning=FALSE}
# From Euler.R
## Modifications by Dylan
ExponentialGrowth<-function(x,lambda){
  return (lambda*x)
}

Euler<-function(x, D, FUN,lambda){
  return(x+FUN(x,lambda)*D)
}

ExponentialSolution<-function(N0,t,lambda){
  return (N0*exp(t*lambda))
}

N0<-0.25
lambda<-0.3
errorsDF <- data.frame(dt = c(0.05,0.1,0.2), Error = NA) # init df
plot(c(0,0),col="0",xlim=c(0,10),ylim=c(0,(N0*exp(lambda*10)+1)))
for (Deltat in c(0.05,0.1,0.2)){
  PointsToEstimate<-seq(0,10,by=Deltat)
  Iterations<-length(PointsToEstimate)-1
  Approx<-rep(0,Iterations+1)
  RealSol<-rep(0,Iterations+1)
  Approx[1]<-RealSol[1]<-N0
  for (i in 1:Iterations){
    RealSol[i+1]<-ExponentialSolution(N0,Deltat*i,lambda)		
    Approx[i+1]<-Euler(Approx[i],Deltat,ExponentialGrowth,lambda)		
  }
  if (Deltat==0.05) {
    mycol="red"
    errorsDF[1,2] <- abs(RealSol[length(RealSol)] - Approx[length(Approx)])
    # last element in approx/real vectors is at time = 10
    # update data frame with error at t = 10 for each Deltat
  }
  if (Deltat==0.1) {
    mycol="blue"
    errorsDF[2,2] <- abs(RealSol[length(RealSol)] - Approx[length(Approx)])
  }
  if (Deltat==0.2)  {
    mycol="green"
    errorsDF[3,2] <- abs(RealSol[length(RealSol)] - Approx[length(Approx)])
  }
  
  points(RealSol~PointsToEstimate,col="black",type="l")
  points(Approx~PointsToEstimate,col=mycol,type="l")
}

ggplot(errorsDF, aes(errorsDF$dt, errorsDF$Error)) +
  geom_line() +
  labs(x = 'Step Size', y = 'Error at t = 10', title = 'Error as fxn of step size') +
  theme(
    plot.title = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )
```
Euler's Method error at $\Delta{t} = (0.005, 0.1, 0.2)$ for exponential growth. Error values were obtained from the difference between the absolute values of the analytic solution and the numerical solution. Error increases rapidly as step size increases.

# Problem 5
```{r results='hide', message=FALSE, warning=FALSE}
## ExpGrowth.R 
## Stefano Allesina sallesina@uchicago.edu allesinalab.uchicago.edu
## Code for "Theoretical Ecology"
## Modifications by Dylan

require(deSolve) 

## This function takes a step in time
ExponentialGrowth<-function(t, state, parameters) { 
  X <- state[1] ## the first element is the density of X at time t
  r <- parameters[1] ## the first parameter is the growth rate
  dX<-r*X 
  return(list(dX)) ## for some reason, you have to return a list
}

## This function runs the model and produces the trajectory
RunExponentialGrowth<-function(MaxTime=10,GrowthRate=0.1,InitialX=1.0){
  times <- seq(0, MaxTime, by = 0.01)
  parameters <- c(r=GrowthRate)
  state <- c(X=InitialX)
  out <- ode(y = state, times = times, func = ExponentialGrowth, parms = parameters)
  print(plot(out[,2]~out[,1],type="l",xlab="time",ylab="Density of X"))
  return(out)
}

logisticGrowth <-function(t, state, parameters) { 
  X <- state[1] ## the first element is the density of X at time t
  r <- parameters[1] ## the first parameter is the growth rate
  dX<-r*X*(1-(X/r))
  return(list(dX)) 
}

runLogisticGrowth<-function(MaxTime=10,GrowthRate=0.1,InitialX=1.0){
  times <- seq(0, MaxTime, by = 0.01)
  parameters <- c(r=GrowthRate)
  state <- c(X=InitialX)
  out <- ode(y = state, times = times, func = logisticGrowth, parms = parameters)
  print(plot(out[,2]~out[,1],type="l", xlab="time",ylab="Density of X"))
  return(out)
}

alleeGrowth<-function(t, state, parameters) { 
  X <- state[1]  
  r <- parameters[1] ## the first parameter is the growth rate
  dX<- (((r * X) / (5 + X)) - (0.1 * X)) * X # parameters hard-coded from previous Allee exercise
  return(list(dX)) 
}

runAlleeGrowth<-function(MaxTime=10,GrowthRate=0.1,InitialX=1.0){
  times <- seq(0, MaxTime, by = 0.01)
  parameters <- c(r=GrowthRate)
  state <- c(X=InitialX)
  out <- ode(y = state, times = times, func = alleeGrowth, parms = parameters)
  print(plot(out[,2]~out[,1],type="l", xlab="time",ylab="Density of X"))
  return(out)
}

levinsGrowth<-function(t, state, parameters) { 
  X <- state[1] 
  r <- parameters[1] 
  dX<- 0.1 * X * (1-X) - exp(X) # c parameter taken from previous Allee exercise
  return(list(dX)) 
}

runLevinsGrowth <-function(MaxTime=10,GrowthRate=0.1,InitialX=1.0){
  times <- seq(0, MaxTime, by = 0.01)
  parameters <- c(r=GrowthRate)
  state <- c(X=InitialX)
  out <- ode(y = state, times = times, func = ExponentialGrowth, parms = parameters)
  print(plot(out[,2]~out[,1],type="l",xlab="time",ylab="Density of X"))
  return(out)
}

thetaLogisticGrowth<-function(t, state, parameters) { 
  X <- state[1] 
  r <- parameters[1] 
  dX<- 2 * X * (1 - (X/15)**1.5) # using theta given on pg. 14, k from exercise 2
  return(list(dX))
}

runThetaLogisticGrowth <-function(MaxTime=10,GrowthRate=0.1,InitialX=1.0){
  times <- seq(0, MaxTime, by = 0.01)
  parameters <- c(r=GrowthRate)
  state <- c(X=InitialX)
  out <- ode(y = state, times = times, func = thetaLogisticGrowth, parms = parameters)
  print(plot(out[,2]~out[,1],type="l",xlab="time",ylab="Density of X"))
  return(out)
}

RunExponentialGrowth(MaxTime=10,GrowthRate=0.1,InitialX=1.0)
runLogisticGrowth(MaxTime=10,GrowthRate=0.1,InitialX=1.0)
runAlleeGrowth(MaxTime=10,GrowthRate=0.1,InitialX=1.0)
runLevinsGrowth(MaxTime=10,GrowthRate=0.1,InitialX=1.0)
runThetaLogisticGrowth(MaxTime=10,GrowthRate=0.1,InitialX=1.0)
```
Numerical solutions for the logistic, Allee, Levins, and theta-logistic ODEs over time.