---
title: "Theo Eco HW 3"
author: "Dylan Sosa"
date: "1/31/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)
```

# Problem 1

1. Solve the SIR models numerically for different diseases. Produce plots for the three classes (S,I,R) over time for typical numbers for Influenza: $R_0 = 2.5, \gamma=\frac{1}{7}$ and HIV: $R_0 = 4, \gamma=\frac{1}{10}$. Check that $S(R)=S_0e^{-R_0\frac{R}{N}}$ is satisfied by the solutions in both cases by plotting S given by this function of R (which you obtain from the numerical solution) along side the solutions from the numerical integration above. 
		
```{r, message=FALSE, warning=FALSE}
SIR.model <- function(t, b, g){
	require(deSolve)
	init <- c(S=1-1e-6,I=1e-6,R=0) 
	parameters <- c(bet=b,gamm=g) 
	time <- seq(0,t,by=t/(2*length(1:t))) 
	
	eqn <- function(time,state,parameters){ 
	with(as.list(c(state,parameters)),{ 
		dS <- -bet*S*I 
		dI <- bet*S*I-gamm*I 
		dR <- gamm*I 
	  return(list(c(dS,dI,dR)))})
	}
	
	out<-ode(y=init,times=time,eqn,parms=parameters) 
	out.df<-as.data.frame(out)
}
```

## Influenza S,I,R Plots
$\beta=R_0\gamma=2.5(\frac{1}{7})\approx0.36$
```{r}
influenzaOutput <- SIR.model(500, 0.36, 1/7)
influenzaOutput$SR <- (1- 1e-6)* exp(-2.5*influenzaOutput$R)

ggplot(influenzaOutput) +
  geom_line(aes(influenzaOutput$time, influenzaOutput$S, color = 'S'), show.legend = TRUE) +
  geom_line(aes(influenzaOutput$time, influenzaOutput$I, color = 'I'), show.legend = TRUE) +
  geom_line(aes(influenzaOutput$time, influenzaOutput$R, color = 'R'), show.legend = TRUE) +
  geom_line(aes(influenzaOutput$time, influenzaOutput$SR,  color = 'S given by f(R)'), alpha = .7, show.legend = TRUE) +
  xlab('Days') +
  ylab('Proportion') +
  ggtitle('Influenza S,I,R Plots') +
  scale_colour_manual("Legend", values = c("forestgreen", "brown", 'blue', 'yellow')) +
  theme(
    plot.title = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )

```

## HIV S,I,R, Plots
$\beta=R_0\gamma=4(\frac{1}{10})=0.4$
```{r}
hivOutput <- SIR.model(500, 0.4, 1/10)
hivOutput$SR <- (1- 1e-6)* exp(-2.5*hivOutput$R)

ggplot(hivOutput) +
  geom_line(aes(hivOutput$time, hivOutput$S, color = 'S'), show.legend = TRUE) +
  geom_line(aes(hivOutput$time, hivOutput$I, color = 'I'), show.legend = TRUE) +
  geom_line(aes(hivOutput$time, hivOutput$R, color = 'R'), show.legend = TRUE) +
  geom_line(aes(hivOutput$time, hivOutput$SR,  color = 'S given by f(R)'), alpha = .7, show.legend = TRUE) +
  xlab('Days') +
  ylab('Proportion') +
  ggtitle('HIV S,I,R Plots') +
  scale_colour_manual("Legend", values = c("forestgreen", "brown", 'blue', 'yellow')) +
  theme(
    plot.title = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )

```


# Problem 2

2. Numerical Integration with Different Solvers. Run the model for 30 days with N = 100, 000. How many people have recovered after 5 days? How many after 10? Plot the curves of S, I, and R

```{r, message=FALSE}
SIR.RKmodel <- function(t, r, g){
	require(deSolve)
	init <- c(R=0) 
	parameters <- c(rnot=r,gamm=g) 
	time <- seq(0,t,by=t/(2*length(1:t))) 
	eqn <- function(time,state,parameters){ 
	with(as.list(c(state,parameters)),{ 
		dR <- gamm*(1-(0.9999*exp(-rnot*R)+R))
	  return(list(c(dR)))})
	}
	out<-rk4(y=init,times=time,eqn,parms=parameters) # use runge-kutta 4th order method
	out.df<-as.data.frame(out)
}

measlesOutput <- SIR.RKmodel(30, 18, 1/5)
measlesOutput$S <- (0.9999)*exp(-18*measlesOutput$R)
measlesOutput$I <- 1-measlesOutput$R-measlesOutput$S

print(
  paste('People recovered after 10 days:',
  as.numeric(filter(measlesOutput, time == 10) %>% 
  select(R)*1e5))
  )

print(
  paste('People recovered after 5 days:',
  as.numeric(filter(measlesOutput, time == 5) %>% 
  select(R)*1e5))
  )


ggplot(measlesOutput) +
  geom_line(aes(measlesOutput$time, measlesOutput$S, color = 'S'), show.legend = TRUE) +
  geom_line(aes(measlesOutput$time, measlesOutput$I, color = 'I'), show.legend = TRUE) +
  geom_line(aes(measlesOutput$time, measlesOutput$R, color = 'R'), show.legend = TRUE) +
  xlab('Days') +
  ylab('Proportion') +
  ggtitle('Measles S,I,R Plots') +
  scale_colour_manual("Legend", values = c("forestgreen", "brown", 'blue')) +
  theme(
    plot.title = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )
```

# Problem 3
3. $R_0$ and competitive exclusion. Using invasion analysis, we showed that for one pathogen to invade another in a SIR model with demography and no co-infection, it needs to have a higher $R_0$. *We could also formally show that when there are two competing pathogens with different $R_0$ values in this model, the one with the greater $R_0$ will outcompete the other.* How is this result consistent with competitive exclusion (remember the cows and the bunnies)? Use mathematical notation and explain in terms of the resource use of each species.

A new pathogen can invade a population and overtake the endemic pathogen if $R'_0 > R_0$. In the competitive exclusion example with cows and bunnies, there is unstable equilibrium when they compete for the same resources. 

